\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-03 main.tex
Fix latex errors
Assets and code listing
Wide images
Recent changes

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{tabularx}

@@

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

@@

\begin{abstract}
In recent years, many programming communities have been undergoing a transition to writing strongly typed programs. To benefit from type-safe codebases, programmers must first overcome the challenge of resolving type errors where text messages can often be unhelpful and misleading. Modern programming tools do not concentrate on making type errors easy to understand. We demonstrate \chameleon{}, a type debugging tool capable of displaying the full context of a type error and allowing programmers to search for where the type error actually occurs based on their prior knowledge and to incrementally verify their assumptions. \chameleon{} was developed through a user-centred, iterative process with four participatory experiments. Through the experiments, we showed that programmers using \chameleon{} fix type errors faster than traditional text-based error messages.  This difference is more significant in harder tasks. Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.
\end{abstract}

@@

\bibliographystyle{IEEEtran}
\bibliography{reference}

>>>>

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{tabularx}
\usepackage{minted}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}

\usemintedstyle{borland}

@@

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\newcommand{\ignore}[1]{}

\begin{document}

@@

\ignore{
\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}
}

@@

\begin{abstract}
In recent years, many programming communities have been undergoing a transition to writing strongly typed programs. To benefit from type-safe codebases, programmers must first overcome the challenge of resolving type errors where text messages can often be unhelpful and misleading. Modern programming tools do not concentrate on making type errors easy to understand. We demonstrate \chameleon{}, a type debugging tool capable of displaying the full context of a type error and allowing programmers to search for where the type error actually occurs based on their prior knowledge and to incrementally verify their assumptions. \chameleon{} was developed through a user-centered, iterative process with four participatory experiments. Through the experiments, we showed that programmers using \chameleon{} fix type errors faster than traditional text-based error messages.  This difference is more significant in harder tasks. Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.
\end{abstract}

@@

\bibliographystyle{IEEEtran}
\bibliography{icpc}

### 2022-12-03 sections/chameleon.tex
Wide images
Recent changes

\chameleon{} comprises two parts: a type inference engine and an original debugging interface. This separate architecture allows us to 1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications, and 2) reuse the debugging interface for different programming languages by replacing the back end type inference engine. The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements.

@@

Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting and introduced a design for interactive type debugging for the Haskell programming language. Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ghc}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving. In Chameleon, Constraints are generated from the source code based on typing rules. In addition, each constraint is labeled with the location (line and column numbers) where it is generated. This set of constraints is consistent if the program is well-typed, inconsistent if otherwise. When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contains inconsistencies. This subset is called a Minimal Unsatisfiable Subset (MUS). From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS. Stuckey et al.~\cite{stuckey2003interactive} showed that program locations linked to the constraints from a MUS are all relevant to the type error and must include the cause of the error. In this paper, we will refer to these locations as \textit{error locations}.

Despite successfully borrowing the underlying ideas, we could not reuse the implementation of Chameleon due to the project language standard and libraries used was out of date. In addition to what Chameleon can do, a few new type inference features were added in the \chameleon{} implementation.

@@

Using only constraints from the MUS is sufficient to locating the type error, but to recover types from type errors we need the constraints from parts of the program that irrelevant to the type errors.  For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}. The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}. Although the recovered types are theoretically correct, they introduce new type variable \texttt{a} making the error message harder to understand.  To solve this issue in \chameleon{}, we first expand the MUS to a maximally satisfiable subset. We do so by removing one constraint from the MUS to get a consistent set, and then repeatedly adding constraints from the global constraint store while maintaining consistency. This process stops when no more constraints can be added. The resulting subset, while not helpful to inform error locations, will produce the most concrete conflicting types. This difference can be seen in figure~\ref{fig:compare-to-original}.

@@

In addition, \chameleon{} provides support for type explanation. Similar to the type explanation system ~\cite{jun2002explaining} from Yang,  \chameleon{} is able to produce a human-readable explanation, but for type errors. This is achieved by expanding the metadata fields of each constraint to contain the information needed for type explanation.

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}

\begin{minted}{haskell}
if a then b else c
a = "True"
\end{minted}

For instance, the program in the listing \ref{listing:ifelse} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS, for example: \texttt{a} has type \texttt{Bool}, because \texttt{a} is the condition of an if statement; however \texttt{a} has type \texttt{String}, because \texttt{a} is defined as the string literal \texttt{"True"}. This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/atonomy.pdf}
    \caption{
        \textbf{The anatomy of \chameleon{}.}
        \chameleon{} consists of two windows: an editor window on the left and a
        debugging window on the right. The editor window is similar to a
        traditional code editor. Fragments of source code may have a highlight
        color (A). Additionally, an explanation layer (B) displays in the editor
        window, if deduction steps are enabled. The debugging window contains
        three panels. A message panel contains an error description (C), and, optionally,
        a list of candidate expression cards (D), a list of deduction steps (E), 
        and a control bar (F) to increment/decrement deduction step. 
        The conflicting types
        panel (G) shows two possible types inferred by the type inference
        engine. A relevant type information panel to show additional information
        that may help understand type errors.
    }
    \label{fig:anatomy}
\end{figure}

@@

The \chameleon{} debugging interface provides three main features to visualize the relation between an error statement and the error locations in the source code, and to explore the explanation of each error location given by the type inference engine. Further, programmers  can turn on and off features to suit their preferences and debugging needs.

\paragraph{Type compare tool} \label{sub:type-compare}

The type compare tool shows the two conflicting types in different colors, each associated with one or more error locations (Fig. \ref{fig:compare}).  The type error must be fixed by modifying at least one of these locations. The error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. To make the bisecting effect more pronounced, a user can hover on one of the possible types and only show the relevant locations that
contribute to that type. This is a convenient way to put the error "under the spotlight".

@@

A candidate expression is an expression that can be inferred to have two conflicting types. 
We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict. By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" here x is the candidate expression. In practice this candidate expression often does not match programmers' intention (Fig. \ref{fig:single-candidate}).

@@

This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third party libraries. With candidate expression cards, a programmer can freely change the context of the type error to the user defined variables and functions to gain a better understanding of why her own code does not match the library code instead of the other direction.

@@

Deduction steps are motivated by the lack of explain why the program has  type errors. When compiler reject a program, it dumps the internal state of type checking.  This information may be a result of complex computing but this process is not reported by compilers. For a typical type error shown in Fig. \ref{fig:ghc-error-example}, clearly the evidence for type error are gathered from are generated from the previous two declarations. These have to be re-discovered by the programmers again, using harder and less sound methods.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/step-interface.pdf}
    \caption{
Deduction steps if they are shown all at once (left). In practice, steps are shown one at a time. Programmers increment or decrement the step number using the step control bar (Fig. \ref{fig:anatomy}-F) or by directly clicking on a step button (Fig. \ref{fig:anatomy}-E). To increment or decrement the deduction step can be intuitively thought of as to move the position of the \textit{split point} where the blue and orange highlights divide (right).
        }
    \label{fig:step-interface}
\end{figure}

@@

Nielson pointed out \cite{nielsen1994usability}  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users. From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task. It is common that a programmer wants the debugger to just "show the answer", and it is also possible a programmer wants to dive deeper into the problem domain and search for the optimal solution. To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced. Programmers can switch between modes by clicking on the mode switching toggles (Fig. \ref{fig:anatomy}-I). The features accessible from different modes are shown in table~\ref{tab:chameleon-features}.

>>>>

\chameleon{} comprises two parts: a type inference engine and an original debugging interface. This separate architecture allows us to 1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications, and 2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine. The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements.

@@

Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting and introduced a design for interactive type debugging for the Haskell programming language. Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving. In Chameleon, Constraints are generated from the source code based on typing rules. In addition, each constraint is labeled with the location (line and column numbers) where it is generated. This set of constraints is consistent if the program is well-typed and inconsistent otherwise. When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contain inconsistencies. This subset is called a Minimal Unsatisfiable Subset (MUS). From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS. Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from a MUS are all relevant to the type error and must include the cause of the error. In this paper, we will refer to these locations as \textit{error locations}.

Despite successfully borrowing the underlying ideas, we could not reuse the implementation of Chameleon since the project language standard and libraries used were out of date. In addition to what Chameleon can do, a few new type inference features were added in the \chameleon{} implementation.

@@

Using only constraints from the MUS is sufficient to locate the type error, but to recover types from type errors we need  constraints from parts of the program that are irrelevant to the type error.  For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}. The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}. Although the recovered types are theoretically correct, they introduce a new type variable \texttt{a} making the error message harder to understand.  To solve this issue in \chameleon{}, we first expand the MUS to a maximally satisfiable subset. We do so by removing one constraint from the MUS to get a consistent set, and then repeatedly adding constraints from the global constraint store while maintaining consistency. This process stops when no more constraints can be added. The resulting subset, while not helpful to inform error locations, will produce the most concrete conflicting types. This difference can be seen in figure~\ref{fig:compare-to-original}.

@@

In addition, \chameleon{} provides support for type explanation. Similar to the type explanation system ~\cite{jun_explaining_2002} from Yang,  \chameleon{} is able to produce a human-readable explanation, but for type errors. This is achieved by expanding the metadata fields of each constraint to contain the information needed for type explanation.

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}
\begin{listing}[!ht]
\begin{minted}{haskell}
if a then b else c
a = "True"
\end{minted}
\caption{A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.}
\label{listing:1}
\end{listing}
For instance, the program in the listing \ref{listing:1} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS, for example, \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}. This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{images/atonomy.pdf}
    \caption{
        \textbf{The anatomy of \chameleon{}.}
        \chameleon{} consists of two panes: an editor pane on the left and a
        debugging pane on the right. The editor pane is similar to a
        traditional code editor. Fragments of source code may have a highlight
        color (A). Additionally, an explanation layer (B) displays in the editor
        window, if deduction steps are enabled. The debugging window contains three blocks. 
        An error statement and configuration block, containing
        An error statement (D), and, optionally,
        a list of candidate expression cards (E), a list of deduction steps (F), 
        and a control bar (G) to increment/decrement deduction step. 
        The conflicting types
        block shows two alternative types (H) inferred by the type inference
        engine. A relevant type information block to show additional information (I)
        that may help understand type errors.
    }
    \label{fig:anatomy}
\end{figure*}

@@

The \chameleon{} debugging interface provides three main features to visualize the relation between an error statement and the error locations in the source code and to explore the explanation of each error location given by the type inference engine. Further, programmers  can turn on and off features to suit their preferences and debugging needs.

\paragraph{Type compare tool} \label{sub:type-compare}
In designing \chameleon{}, we want to argue against the practice of reporting type errors as the conflicts between \texttt{Expected type} and \texttt{Actual type}. Technically, the two alternatives are merely the side effect of different unification orders of the internal type-checking technique and have no bearing on what the programmers think and expect.

The type compare tool shows the two conflicting types in different colors, each associated with one or more error locations (Fig. \ref{fig:compare}).  The type error must be fixed by modifying at least one of these locations. The error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. To make the bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that
contribute to that type. This is a convenient way to put the error \texttt{under the spotlight}.

@@

A candidate expression is an expression that can be inferred to have two conflicting types. 
We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict. By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression. In practice, this candidate expression often does not match programmers' intention (Fig. \ref{fig:single-candidate}).

@@

This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third-party libraries. With candidate expression cards, a programmer can freely change the context of the type error to the user-defined variables and functions to gain a better understanding of why her own code does not match the library code instead of the other direction.

@@

Deduction steps are motivated by the lack of explaining why the program has type errors. When the compiler rejects a program, it dumps the internal state of type checking.  This information may be a result of complex computing but this process is not reported by compilers. For a typical type error shown in Fig. \ref{fig:ghc-error-example}, clearly, the evidence for the type error is gathered from the previous two declarations. These have to be rediscovered by the programmers again, using harder and less-sound methods.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/step-interface.pdf}
    \caption{
Deduction steps if they are shown all at once (left). In practice, steps are shown one at a time. Programmers increment or decrement the step number using the step control bar (Fig. \ref{fig:anatomy}-G) or by directly clicking on a step button (Fig. \ref{fig:anatomy}-F). To increment or decrement the deduction step can be intuitively thought of as moving the position of the \textit{split point} where the blue and orange highlights divide (right).
        }
    \label{fig:step-interface}
\end{figure}

@@

Nielson pointed out \cite{jakob_nielsen_usability_1993}  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users. From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task. It is common that a programmer wants the debugger to just "show the answer", and it is also possible a programmer wants to dive deeper into the problem domain and search for the optimal solution. To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced. Programmers can switch between modes by clicking on the mode switching toggles (Fig. \ref{fig:anatomy}-C). The features accessible from different modes are shown in table~\ref{tab:chameleon-features}.

### 2022-12-03 sections/conclusion.tex
Recent changes

We present \chameleon{}, a type debugging tool for the Haskell programming language. \chameleon{} facilitates a constraint-based type inference back end that provides accurate and comprehensive type error reports. \chameleon{} also facilitates an interactive front end with three different display modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code. Four participatory experiments were conducted to evaluate the tool's design with programmers with varied experiences. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages.  Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.

>>>>

We present \chameleon{}, a type debugging tool for the Haskell programming language. \chameleon{} facilitates a constraint-based type inference engine that provides accurate and comprehensive type error reports. \chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code. Four experiments with human participants were conducted to evaluate the tool's design with programmers with varied experiences. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages.  Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.

### 2022-12-03 sections/discussion.tex
Assets and code listing
Recent changes

From the results of user studies 1 and 2, we observed that the choice of debugging tool had little effect on how fast programmers solve type errors. Conversely, when facing more realistic problems (longer source code, error locations more scattered), programmers are more effective using \chameleon{}. We believe one factor of this effectiveness of \chameleon{} is to reduce the amount of reading time. Jbara \cite{jbara2017programmers} showed that reading source code is generally the initial step of solving programming problems, and is done in several passes. Although traditional compiler error message tools initially show fewer locations, it is unreliable, meaning that programmers have to expand the reading span without clear guidance. In contrast, Chameleon shows more error locations initially. However, the completeness of error locations assures programmers which part of the source code can be safely skipped.

@@

\begin{lstlisting}[language=Haskell, float, caption={
    In this simple program, \chameleon{} report 
    the error happened in \texttt{f} and \texttt{z}.
}, label={listing:fz}]
f z     
    | z == 3 = False
    | z == '4' = True
\end{lstlisting}

We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program showing in the listing \ref{listing:fz}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although, in theory, these two statements are to the same effect, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is preserved on screen and can be retrieved anytime. This may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

@@

Another area for future study is to extend on the evaluation process. In the user studies 2, 3, and 4, we used harder tasks compared to the ones in user study 1. However, the difficulty level is nowhere as hard as a professional Haskell programmer may face in the production Haskell codebase. It would be interesting to see how \chameleon{}  faces type errors that span multiple files and packages. Another dimension of difficulty is higher-level extractions. Haskell programs get more tricky to the debugger when more abstractions are added to the codebase, such as Monads, Monad transformers, and Lenses. These topics were not included in our study to accommodate novice users. However, observing how expert Haskell programmers use \chameleon{} features to solve advanced problems will gain valuable insight into the true usability of the tool.

@@

Our study focuses on the Haskell language for its popularity in the academic world. However, the low-quality of textual type errors is a problem not limited to Haskell. Modern statically typed languages more or less all share the same problem. We believe the underlying type reporting technique and algorithms can be generalized to other languages. It will be exciting to see how interactive debugging features perform in other paradigms, such as  imperative languages and incremental type systems.

Our design of \chameleon{}  emphasizes allowing programmers to update the type error based on their prior knowledge. It unfolds to answering three questions: which expression should be considered un-typeable? Between multiple possible types, which one should be considered intended? Among all the possible locations, which one should be considered the cause? \chameleon{} answers all three questions by leaving it to the programmers to decide. However, it is possible and likely that some programmers enjoy conclusive error messages instead of flexible ones provided by \chameleon{}. Using heuristic methods may achieve the best of both worlds. It may produce effective results by guessing the possible un-typable expression based on the position it appears in the syntax tree and project structure or guessing the intended type signature by comparing the number of supporting slices in the source code.

>>>>

From the results of user studies 1 and 2, we observed that the choice of debugging tool had little effect on how fast programmers solve type errors. Conversely, when facing more realistic problems (longer source code, error locations more scattered), programmers are more effective using \chameleon{}. We believe one factor of this effectiveness of \chameleon{} is to reduce the amount of reading time. Jbara \cite{jbara2017programmers} showed that reading source code is generally the initial step of solving programming problems and is done in several passes. Although traditional compiler error message tools initially show fewer locations, it is unreliable, meaning that programmers have to expand the reading span without clear guidance. In contrast, Chameleon shows more error locations initially. However, the completeness of error locations assures programmers which part of the source code can be safely skipped.

@@

\begin{listing}[!ht]
\begin{minted}{haskell}
f z     
    | z == 3 = False
    | z == '4' = True
\end{minted}
\caption{     In this simple program, \chameleon{} report the error happened in \texttt{f} and \texttt{z}.}
\label{listing:2}
\end{listing}

We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program showing in the listing \ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although, in theory, these two statements are to the same effect, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime. A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

@@

Another area for future study is to extend the evaluation process. In studies 1b and 2, we used harder tasks compared to study 1a. However, the difficulty level is nowhere as hard as a professional Haskell programmer may face in the production Haskell codebase. It would be interesting to see how \chameleon{}  faces type errors that span multiple files and packages. Another dimension of difficulty is higher-level extractions. Haskell programs get more tricky to the debugger when more abstractions are added to the codebase, such as Monads, Monad transformers, and Lenses. These topics were not included in our study to accommodate novice users. However, observing how expert Haskell programmers use \chameleon{} features to solve advanced problems will gain valuable insight into the true usability of the tool.

@@

Our study focuses on the Haskell language for its popularity in the academic world. However, the low quality of textual type errors is a problem not limited to Haskell. Modern statically typed languages more or less all share the same problem. We believe the underlying type reporting technique and algorithms can be generalized to other languages. It will be exciting to see how interactive debugging features perform in other paradigms, such as  imperative languages and incremental type systems.

Our design of \chameleon{}  emphasizes allowing programmers to update the type error based on their prior knowledge. It unfolds to answer three questions: which expression should be considered un-typeable? Between multiple possible types, which one should be considered intended? Among all the possible locations, which one should be considered the cause? \chameleon{} answers all three questions by leaving it to the programmers to decide. However, it is possible and likely that some programmers enjoy conclusive error messages instead of flexible ones provided by \chameleon{}. Using heuristic methods may achieve the best of both worlds. It may produce effective results by guessing the possible un-typable expression based on the position it appears in the syntax tree and project structure or guessing the intended type signature by comparing the number of supporting slices in the source code.
